---
title: "CS 422 HW1"
output:
  html_notebook:
    toc: yes
    toc_float: yes
author: "Javier Echavarren Suárez"
---

### Part i

```{r}
data.df<-read.csv(file = "us-covid.csv", skip =  7)

```

### Part ii

```{r}
data.df = subset(data.df, select = -c(confirmed_deaths) )

```

### Part iii - a

```{r}
library(dplyr)
arrange(data.df, desc(total_cases))
arrange(top_n(data.df,6,total_cases), desc(total_cases))

```
### Part iii - b
```{r}
library(dplyr)
arrange(data.df, desc(total_cases))
arrange(top_n(data.df,-6,total_cases), desc(total_cases))

```

### Part iv

```{r}
library(psych)
pairs.panels(subset(data.df, select = -c(state, level) ))

```

#### Part iv a
(a) Why should we exclude the first two columns from correlation analysis?

Because they are nominal data, and we can't make correlation with this type of data

#### Part iv b
(b) Which pair of columns have the highest correlation?

Total cases and total deaths, with a correlation of 0.98

#### Part iv c
(c) Which pair of columns has the lowest correlation?

Cases in the last 7 days and case rate per 100k

### Part v

```{r}
data.df<-mutate(data.df, confirmed_plus_probable = confirmed + probable)
plot(data.df$confirmed_plus_probable,data2$total_deaths , xlab = "Confirmed + probable" , ylab = "Total deaths")

```
#### Part v b
(b) As you see the plot, there appears to be an anomaly. Looking at the data, briefly describe the anomaly.

The anomaly is that there are states in which confirmed + probable cases are 0, and that is because we lack the data of those states

#### Part v c

```{r}
select(filter(data.df , confirmed_plus_probable == 0), state,total_deaths, confirmed , probable , total_cases)


```
